Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-19 src/tex/ms.tex
[showyourwork] overleaf sync

\paragraph{Posteriors for Galactic parameters}
As stated in \citet{Philcox_2019} the full posterior for this case is given by 
\begin{eqnarray}\label{eq:posterior}
    \mathbb{P}(\vec\Lambda,\{\vec\Theta_i\},\{T_i\}|\{D_i\}) &\propto&  \left[\prod_{i=1}^{n_\mathrm{star}}p_{\vec\Theta}(\vec\Theta_i)p_{T_i}(T_i)\right]
    \times p_{\vec\Lambda}(\vec\Lambda)\\
    \nonumber
    &\times& \mathcal{L}(\{D_i\}|\vec\Lambda,\{\vec\Theta_i\},\{T_i\})
\end{eqnarray}
where $p_V(V_i)$ is the prior on variable $V_i$ (belonging to the set $V$).

@@

\paragraph{Simulation-based inference}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a Bayesian inference framework -- works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\Vec{\theta}$ (in our case a GCE model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\Vec{\theta}|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\Vec{\theta}$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to observations of stellar abundances $\Vec{X_R}$ to infer $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$) (see Fig. \ref{fig:flowchart} for a schematic visual representation of our method). Note, we do not need to know anything about $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$), we solely need to be able to sample from it.

We use a NPE \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow (NSF) \citet{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value \citep{Lemos2023}. For more details on the TARP value see section \ref{sec:sbc} in the appendix.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section with inputs sampled from a uniform prior over the original $5\sigma$-range of the Gaussian priors shown in Tab. \ref{tab:priors} to provide better coverage of the parameter space. Training takes $\sim70$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPU cores. On an Apple M1 Max training only takes 36 minutes.
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

Note, the methods presented here can naturally be extended to any fast and flexible GCE model, not just \texttt{CHEMPY}.

\paragraph{Multi-star inference}
Following eq.~\ref{eq:posterior} we can calculate the joint posterior for a combined inference using abundance observations of multiple stars. For this, we will condition our NPE individually on single star observations and sample the posterior. Then, according to eq.~\ref{eq:posterior} we simply have to multiply individual posteriors and account for the inverse weighting by some power of the prior which we take from table~\ref{tab:priors}.

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate eq.~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

\subsection{\texttt{CHEMPY} Mock Observational Data}
Our analysis uses mock observations drawn from the neural network emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and varying local parameters $\vec{\Theta}_i$ using priors from Tab. \ref{tab:priors}. Additionally we draw $T_i$ uniformly in the range $[2,12.8]$ Gyr to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to our \texttt{CHEMPY} emulator, producing eight true chemical abundances. In order to fully represent observational data, we augment this data with observational uncertainties by adding a Gaussian error of $0.05$ dex for the abundances representative for typical APOGEE data \citep{Majewski2016}.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainties as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions). See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield} for a full description.}
    \label{fig:CHEMPY_alt_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data taken from an IllustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim} for a full description.}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

@@

\paragraph{Validation through absolute percentage error}
We start evaluating our method by using \texttt{CHEMPY} to produce a mock observational dataset to ensure no systematical shift between training and testing data in terms of physics parameters.

@@

\subsection{Inference on mock data from \texttt{CHEMPY} with TNG yield set}
\label{subsec:chempy_tng}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1,000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference. See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}

%
Combining the inference on multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$. We therefore perform inference using a range of stars $n\in[1,1000]$ (see Fig. \ref{fig:CHEMPY_TNG_N_star_analysis}) to show how inference accuracy increases with number of observations. We find that in the limit of few stars (less than $\sim100$) SBI shows a larger uncertainties than the HMC results of \citet{Philcox_2019}. But in the limit of large numbers of stars (few hundred) the accuracy and precision is superior compared to HMC. In particular, after using a few tens of stars, the NPE estimate is less biased than the HMC results. 
Finally, given our computational advantage we will be able to use orders of magnitude more stars for our inference.
This is particularly important since sample variances play a large role when using small samples of stars as also noted in \citet{Philcox_2019}.
%

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma$ levels.
%
Using a sample of 1,000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the correct value.

Note that our analysis is in principle also able to infer the local parameters $\Vec{\Theta}_i$ and $T_i$. This would allow us to estimate/infer stellar ages as well. However, currently our NPE network is not particularly good at estimating ages from abundances alone. On average our inference for ages results in an APE of $\sim27\%$ which is slightly larger than the observational noise of $20\%$ that we add during the mock up of our data.

@@

There is an extensive discussion in the literature about stellar nucleosynthesis with various different yield sets proposed \citep[see e.g. discussion in][]{Rybizki_2017}. In fact, any tabulated yield set will differ from reality and during an application of our inference pipeline it will not be clear which tabulated yield set most closely matches reality and hence which should be used.
In order to investigate how sensitive our method is to a model misspecification by using an incorrect yield set, we create another set of mock data using \texttt{CHEMPY} with a different yield set than during training of our NPE. For better cross-comparison we decided to use the same alternative yield sets as presented in Tab.~5 of \citep{Philcox_2019}.

\begin{table}[]
\caption{Alternative nucleosynthetic yield tables used in section \ref{sec: Results} for model misspecification tests.}
     \centering
     \begin{tabular}{c|c}
       Type & Yield Table \\
        \hline
         SN\,Ia & \citet{2003NuPhA.718..139T} \\
         SN\,II & \citet{2013ARA&A..51..457N} \\
         AGB & \citet{2016ApJ...825...26K}
     \end{tabular}
\label{tab:chempy_ALT_yields}
\end{table}

@@

Our mock data generation and inference then follows the one of Sec.~\ref{subsec:chempy_tng}. This means we apply our NPE that we trained on \texttt{CHEMPY} stellar abundances simulated with the Ilustris TNG yield set to a set of stellar abundances simulated with \texttt{CHEMPY} but using the above mentioned alternative yield set. This effectively probes the effect of model missspecification on the inference results.

We show our inference results for this setup for varying number of stars (blue) in Fig.~\ref{fig:CHEMPY_alt_N_star_analysis} comparing again against HMC results (red) from \citet{Philcox_2019}. We see that the SBI results are less biased than the HMC results especially for $\alpha_{\rm IMF}$. In fact with 100 stars HMC inferences are about \textcolor{red}{calculate value} $\sigma$ away from the ground truth value for $\alpha_{\rm IMF}$ and $\log_{10}(N_{\rm Ia})$, respectively. Again, the joint posterior is shown in Fig.~\ref{fig:CHEMPY_alt_sbi} and shows that while individually parameter inferences look good, jointly taken the SBI results are on the edge of being $3\sigma$ away from the ground truth.

Nevertheless, the performance of our inference is still very good. We see that an increasing number of observations helps to decrease the models uncertainty just as before. However, we also note that our inference is now slightly biased as the observational data does not match the training data. Looking at Fig.~\ref{fig:CHEMPY_alt_sbi} we see that our inference is still consistent with the ground truth within the $3\sigma$ levels. Comparing this to the HMC results, we see that SBI is performing much better when the model is misspecified and leads to significantly less biased results. This is next to drastically reduced compute times -- another a key advantage of our SBI method compared to more standard approaches such as HMC.

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 1,000-fold compared to HMC. For instance, while HMC requires $\sim 40\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI’s amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST and WEAVE will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. Even under conditions of mismatched yield tables or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI’s applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}

>>>>

\paragraph{Posteriors for Galactic parameters}
As stated in \citet{Philcox_2019} the full posterior for this case is given by 
\begin{eqnarray}\label{eq:posterior}
    \mathbb{P}(\vec\Lambda,\{\vec\Theta_i\},\{T_i\}|\{D_i\}) &\propto&  \left[\prod_{i=1}^{n_\mathrm{star}}p_{\vec\Theta}(\vec\Theta_i)p_{T_i}(T_i)\right]
    \times p_{\vec\Lambda}(\vec\Lambda)\\
    \nonumber
    &\times& \mathcal{L}(\{D_i\}|\vec\Lambda,\{\vec\Theta_i\},\{T_i\})
\end{eqnarray}
where $p_{\vec\Theta}(\vec\Theta_i)p_{T_i}(T_i)$ are the priors on the variables $\vec\Theta_i$ or $T_i$ belonging to a given set of stars.

@@

\paragraph{Simulation-based inference:}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a Bayesian inference framework -- works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\Vec{\theta}$ (in our case a GCE model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\Vec{\theta}|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\Vec{\theta}$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to observations of stellar abundances $\Vec{X_R}$ to infer $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$). We show a schematic visual representation of our method in Fig. \ref{fig:flowchart}. Note, we do not need to know anything about $p$($\Vec{\theta}|\Vec{X_R}$, $\mathcal{M}$), we solely need to be able to sample from it.

We use a NPE \citep{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow \citep[NSF;][]{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow \citep[MAF;][]{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value \citep{Lemos2023}. For more details on the TARP value see section \ref{sec:sbc} in the appendix.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section. Inputs are sampled from a uniform prior over the original $5\sigma$-range of the Gaussian priors shown in Tab. \ref{tab:priors} to provide better coverage of the parameter space. Training takes $\sim70$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPU cores. On an Apple M1 Max training only takes 36 minutes.
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

Note, the methods presented here can naturally be extended to any fast and flexible GCE model \citep[e.g.][and others]{Talbot1971,vice,omega2018}, not just \texttt{CHEMPY}.

\paragraph{Multi-star inference:}
Following eq.~\ref{eq:posterior} we can calculate the joint posterior for a combined inference using abundance observations of multiple stars. For this, we will condition our NPE individually on single star observations and sample the posterior. Then, according to eq.~\ref{eq:posterior} we simply have to multiply individual posteriors and account for the inverse weighting by some power of the prior which we take from Table~\ref{tab:priors}.

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate eq.~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'^2}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

\subsection{\texttt{CHEMPY} Mock Observational Data}
\label{subsec:mock obs}
Our analysis uses mock observations drawn from the neural network emulator with fixed values of the global galactic parameters $\alpha_{\rm IMF}=-2.3$ and $\log_{10}(N_{\rm Ia})=-2.89$ and varying local parameters $\vec{\Theta}_i$ using priors from Tab. \ref{tab:priors}. Additionally we draw $T_i$ uniformly in the range $[2,12.8]$ Gyr to minimize overlap with the neural network training birth-time limits when observational uncertainties are included. 
Each set of parameters is passed to our \texttt{CHEMPY} emulator, producing eight true chemical abundances. In order to fully represent observational data, we augment this data with observational uncertainties by adding a Gaussian error of $0.05$ dex for the abundances representative for typical APOGEE data \citep{Majewski2016}.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_TNG_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Accuracy of inferred global galactic parameters $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ as a function of number of observed stars, comparing SBI (blue line) to the inferred values using HMC (red line) as done by \cite{Philcox_2019} and the ground truth values (black dashed line). For the SBI analysis we show $1\sigma$ and $2\sigma$ contours while HMC results only show $1\sigma$ statistical uncertainties as reported in Tab.~3 of \citet{Philcox_2019} (blue/red shaded regions). See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_N_star_analysis}
    \script{chempy_tng_inference.py}
\end{figure*}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/CHEMPY_TNG_yields.png}
    \vspace{-.25cm}
    \caption{Joint posterior for global galactic parameters $\alpha_{\rm IMF}$ and $\log_{10}(\rm N_{Ia})$ of $1,000$ stars. The ground truth value is shown by the black dot, the posterior mean of the SBI inference is shown with the red star and white and black ellipses show the $1-3\sigma$ contours of our inference. See Sec.~\ref{subsec:chempy_tng} for a full description.}
    \label{fig:CHEMPY_TNG_sbi} 
    \script{chempy_tng_inference.py}
\end{figure}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/CHEMPY_alternative_yields_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data created with a different yield set than the training data. See Sec.~\ref{subsec:mocks_wrong_yield} for a full description.}
    \label{fig:CHEMPY_alt_N_star_analysis}
    \script{chempy_alt_yield_inference.py}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/TNG_sim_Nstar_comp.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for the mock data taken from an IllustrisTNG Milky Way-like galaxy. See Sec.~\ref{subsec:tng_sim} for a full description.}
    \label{fig:TNG_N_star_analysis}
    \script{tng_inference.py}
\end{figure*}

@@

\paragraph{Validation through absolute percentage error:}
We start evaluating our method by using \texttt{CHEMPY} to produce a mock observational dataset to ensure no systematical shift between training and testing data in terms of physics parameters.

@@

However, currently our NPE network is not particularly good at estimating ages from abundances alone. On average our inference for ages results in an APE of $\sim27\%$ which is slightly larger than the observational noise of $20\%$ that we add during the mock up of our data.

\subsection{Inference on mock data from \texttt{CHEMPY} with TNG yield set}
\label{subsec:chempy_tng}
%
Combining the inference on multiple observed stars gives us higher accuracy and precision of the global galactic parameters $\vec\Lambda$. We therefore perform inference using a range of stars $n\in[1,1000]$ to show how inference accuracy increases with number of observations (see Fig. \ref{fig:CHEMPY_TNG_N_star_analysis}). We find that in the limit of less than $\sim100$ stars SBI shows a larger uncertainties than the HMC results of \citet{Philcox_2019}. But when using more than a few hundred stars the accuracy and precision is superior compared to HMC. In particular, after using a few tens of stars, the NPE estimate is already less biased than the HMC results. 
Finally, given our computational advantage we will be able to use orders of magnitude more stars for our inference.
This is particularly important since sample variances play a large role when using small samples of stars as also noted in \citet{Philcox_2019}.
%

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma$ levels.
%
Using a sample of 1,000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the ground truth value.

Note that our analysis is in principle also able to infer the local parameters $\Vec{\Theta}_i$ and $T_i$. This would allow us to estimate/infer stellar ages as well. But note, as discussed above at the end of sub-section \ref{subsec:mock obs} our NPE currently is not well calibrated to estimate stellar ages accurately enough.

@@

There is an extensive discussion in the literature about stellar nucleosynthesis with various different yield sets proposed \citep[see e.g. discussion in][]{Rybizki_2017}. In fact, any tabulated yield set will differ from reality and during an application of our inference pipeline it will not be clear which tabulated yield set most closely matches reality and hence which should be used.
In order to investigate how sensitive our method is to model misspecification by using an incorrect yield set, we create another set of mock data using \texttt{CHEMPY} with a different yield set (Tab. \ref{tab:chempy_ALT_yields}) than during training of our NPE. For better cross-comparison we decided to use the same alternative yield sets as presented in Tab.~5 of \citep{Philcox_2019}.

\begin{table}[]
\caption{Alternative nucleosynthetic yield tables used for model misspecification tests.}
     \centering
     \begin{tabular}{c|c}
       Type & Yield Table \\
        \hline
         SN\,Ia & \citet{2003NuPhA.718..139T} \\
         %SN\,II & \citet{2013ARA&A..51..457N} \\
         AGB & \citet{2016ApJ...825...26K}
     \end{tabular}
\label{tab:chempy_ALT_yields}
\end{table}

@@

Our mock data generation and inference then follows the one of Sec.~\ref{subsec:chempy_tng}. This means we apply our NPE that we trained on \texttt{CHEMPY} stellar abundances simulated with the Ilustris TNG yield set to a sets of stellar abundances simulated with \texttt{CHEMPY} but using the alternative yield sets mentioned in Tab. \ref{tab:chempy_ALT_yields}. This effectively probes the effect of model misspecification on the inference results.

This isn't up to date
\textcolor{blue}{
In Fig.~\ref{fig:CHEMPY_alt_N_star_analysis} we show our inference results (blue) for this setup for varying number of stars and compare them against HMC results (red) from \citet{Philcox_2019}. We see that the SBI results are less biased than the HMC results, especially for $\alpha_{\rm IMF}$. In fact with 100 stars HMC inferences are about \textcolor{red}{calculate value} $\sigma$ away from the ground truth value for $\alpha_{\rm IMF}$ and $\log_{10}(N_{\rm Ia})$, respectively. Again, the joint posterior is shown in Fig.~\ref{fig:CHEMPY_alt_sbi} and shows that while individually parameter inferences look good, jointly taken the SBI results are on the edge of being $3\sigma$ away from the ground truth.
}

\textcolor{blue}{
Nevertheless, the performance of our inference is still very good. We see that an increasing number of observations helps to decrease the models uncertainty just as before. However, we also note that our inference is now slightly biased as the observational data does not match the training data. Looking at Fig.~\ref{fig:CHEMPY_alt_sbi} we see that our inference is still consistent with the ground truth within the $3\sigma$ levels. Comparing this to the HMC results, we see that SBI is performing much better when the model is misspecified and leads to significantly less biased results. This is next to drastically reduced compute times -- another a key advantage of our SBI method compared to more standard approaches such as HMC.
}

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 1,000-fold compared to HMC. For instance, while HMC requires $\sim 40\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI’s amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST and WEAVE will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. \textcolor{blue}{Even under conditions of mismatched yield tables (should be updated)} or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI’s applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}
